---
title: "Assignment 6"
author: "Xiangling Chen"
date: "2024-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
# Load sf,tmap and tidyverse
library(sf)
library(tmap)
library(tidyverse)
```

## Loading the data
```{r load-data}
## Loading the nc_schools data
nc_schools <- read_sf("https://drive.google.com/uc?export=download&id=1hFPUSj5s-sApJKC5Vh5Ue5wpF_1gbClO")
glimpse(nc_schools)

## Loading nc_county data
nc_counties <-read_sf("https://drive.google.com/uc?export=download&id=1Ks38Gmdlp58GBhOgb6FBmx8tIAfu0QDw")
glimpse(nc_counties)
```

# Exploratory Data Analysis
```{r county-population-eda}
#Summarize the population data looking at descriptive stats
nc_counties|>
  drop_na(popE)|>
  summarize(count =n(),
            "smallest county"=min(popE),
            "25% quantile"=quantile(popE, probs=c(.25)),
            "average size" = mean(popE),
            "median size" = median(popE),
            "75% quantile" = quantile(popE, probs=c(.75)),
            "largest county" = max(popE),
            variance = var(popE))|>
  glimpse()
```

```{r boxplot for popE}
# Making boxplot of population values for NC counties
nc_counties|>
  drop_na(popE)|>
  ggplot(aes(x="popE", y=popE)) +geom_boxplot()
```

```{r counties -population-map}
# Map the counties with the population
map1<-tm_shape(nc_counties)+
  tm_polygons("popE")+
  tm_layout(title = "Population 2020")
map1
```

```{r jenks-classification style}
# Map the counties with the population using the Jenks Classification method
mao1<-tm_shape(nc_counties) + tm_polygons("popE", style ="jenks")+tm_layout(title="Population 2020")
map1
```

```{r counties-popdens-map}
# calculate area and population density for each county
nc_counties<-nc_counties|>
  #calculate the area per county in meter sq
  mutate(area=st_area(nc_counties))|>
  #calculate the population density in kms sq
  mutate(popdens = units::set_units(popE/area,"1/km^2"))
tm_shape(nc_counties)+tm_polygons("popdens",style = "jenks") +tm_layout(title="North Carolina 2020 Populatioin Density")
```

```{r popdens-distribution}
# we need the units library to display the units
library(units)

# make a boxplot of population density for NC counties
nc_counties|>
  drop_na(popdens)|>
  ggplot(aes(y=popdens))+geom_boxplot()
```

```{r county-population-density-eda} 
nc_counties|>
  drop_na(popdens)|>
  summarize(count = n(),
            "smallest county"=min(popdens),
            "25% quantile"=quantile(popdens,probs = c(.25)),
            "average size"=mean(popdens),
            "75% quantile"=quantile(popdens,probs = c(.75)),
            "largest county"=max(popdens),
            variance = var(popdens))|>
  glimpse()
```

```{r counting-missing-values}
# for each column in nc_schools, count the missing values
nc_schools|>
  # select all columns-everything
  select(everything())|>
  # sum the values that are NA(is.na == TRUE) 
  #the ~ says to run the code on each column
  # and the. in is.na(.)substitues the value from the column
  summarize_all(list(~sum(is.na(.))))|>
  glimpse()
```

```{r school-data-eda}
# Summarize the populaiton density looking at descriptive stats
as.data.frame(nc_schools)|>
  select(num_tested_1819:pct_notprof_2122)|>
  summarize_all(.funs=list(min=~min(.,na.rm=TRUE),
                           Q1=~quantile(x=.,probs=c(.25),na.rm=TRUE),
                           mean=~mean(.,na.rm=TRUE),
                           median=~median(.,na.rm=TRUE),
                           Q3=~quantile(x=.,probs=c(.75),na.rm=TRUE),
                           max=~max(.,na.rm=TRUE),
                           var=~var(.,na.rm=TRUE)
                           ))|>
  glimpse()
```

# Boxplots for Variable
```{r boxplots-for-variable}
# make boxplots for the number tested

# this removes the geometry(spatial) data
as.data.frame(nc_schools)|>
  #select only the columns we are about
  select(num_tested_1819:num_tested_2122)|>
  # pivot_longer moves the columns to rows, which make it easier to graph
  pivot_longer(num_tested_1819:num_tested_2122,names_to="names",values_to="values")|>
  #graph it
  ggplot(aes(x=names, y=values))+geom_boxplot()
```



```{r change-outlier}
as.data.frame(nc_schools)|>
  select(pct_notprof_1819:pct_notprof_2122)|>
  pivot_longer(pct_notprof_1819:pct_notprof_2122, names_to="names", values_to="values")|>
  ggplot(aes(x=names, y=values))+geom_boxplot()
```

## Handling Null(NA) Values
```{r replace_nas_with_0}
# replace all null no-data values with 0's
# replace all NA values with 0's in relevant columns
nc_schools.new <- nc_schools |>
  mutate(across(c(num_tested_1819:num_tested_2122, 
                  pct_notprof_1819:pct_notprof_2122), 
                ~replace_na(.x, 0)))
```

```{r pass-rate-per-county}
# calculate, per school, the fail and pass rate
nc_schools.new<-nc_schools.new |>
  mutate(num_fail_1819 = num_tested_1819*pct_notprof_1819/100,
         num_fail_2021 = num_tested_2021*pct_notprof_2021/100,
         num_fail_2122 = num_tested_2122*pct_notprof_2122/100)|>
  mutate(num_pass_1819 = num_tested_1819-num_fail_1819,
         num_pass_2021 = num_tested_2021-num_fail_2021,
         num_pass_2122 = num_tested_2122-num_fail_2122)
# calculate the pass rate per county
# convert the data to data frame to remove geometry
per_county<-as.data.frame(nc_schools.new)|>
  #get county level statistics
  group_by(county)|>
  # compuste the num tested and pass per school
  mutate(num_tested = num_tested_1819+num_tested_2021+num_tested_2122,num_pass = num_pass_1819 + num_pass_2021+num_pass_2122)|>
  #add up the num tested and pass per county
  summarize(across(num_tested:num_pass,sum),numschools=n())|>
  #compute the per conty pass rate
  mutate(pass_pct = num_pass/num_tested *100)
glimpse(per_county)
```

```{r change the name}
glimpse(sort(nc_counties$NAME))
```

```{r join-to-county}
# create a new data frame from nc counties
nc_counties.new<-nc_counties|>
  mutate(name=str_replace(NAME,' County, North Carolina',''))
glimpse(nc_counties.new)
```

```{r join-data}
# join the per_county data and the nc_counties.new.data data together
nc_counties.new<-nc_counties.new |>
  # match name to county
  inner_join(per_county, by=c("name"="county"))
glimpse(nc_counties.new)
```

```{r map the data}
tm_shape(nc_counties.new)+
  tm_polygons("pass_pct")+
  tm_layout(title = "North Carolina 2019-2022 % Student Pass Rate")
```

```{r categorize_popdens}
nc_counties.new<-drop_units(nc_counties.new)
nc_counties.new<-nc_counties.new|>
  mutate(popdens.description = ifelse(popdens<24,"smallest",ifelse(popdens<42,"small",ifelse(popdens<85,"large","largest"))))
glimpse(nc_counties.new$popdens.description)
```

```{r new-box-plot}
nc_counties.new|>
  ggplot(aes(y=pass_pct, x=popdens.description))+geom_boxplot()
```

```{r new-jitter-plot}
# jitter plot show the individual points
nc_counties.new |>
  ggplot(aes(x=popdens.description, y=pass_pct,color=popdens.description)) + geom_jitter()
```

```{r ANOVA test}
# ANOVA looks at the variance
result<-aov(pass_pct~popdens.description, dat=nc_counties.new)
summary(result)
```

## Part 6: Making My own EDA(FemaleE)
```{r femaleE data ploting}
# a)Using tmap to draw the female population in the county
tm_shape(nc_counties.new)+tm_polygons("femaleE",style ="jenks") + tm_layout(title= "females number in NC Counties")
```
```{r ggplot-for the boxplots}
# b) Making boxplot for the values
nc_counties|>
  drop_na(femaleE)|>
  ggplot(aes(x="femaleE",y=femaleE)) + geom_boxplot()
```

```{r printing descriptive variable}
nc_counties.new|>
  drop_na(femaleE)|>
  summarise(count=n(),
            "smallest"=min(femaleE),
            "25% quantile" = quantile(femaleE,probs = (.25)),
            "average size" = mean(femaleE),
            "75% quantile" = quantile(femaleE, probs = c(.75)),
            "largest female number" = max(femaleE),
            variance = var(femaleE))|>
  glimpse()
```

```{r create a group for variable}
nc_counties.new<-drop_units(nc_counties.new)
nc_counties.new <- nc_counties.new |>
  mutate(femaleE.description = ifelse(femaleE < 10000, "small", ifelse(femaleE < 20000, "medium", "large")))
glimpse(nc_counties.new$femaleE.description)
table(nc_counties.new$femaleE.description)
```

```{r making anova test}
nc_counties.new$femaleE.description <- as.factor(nc_counties.new$femaleE.description)
result_2 <- aov(pass_pct ~ femaleE.description, data = nc_counties.new)
summary(result_2)
```

# Part 7 Reflection Question

## 1.How long did the assignment take you to complete?
3hrs

## 2.Where did you get stuck? Where did you go for help?
The categorization for the femaleE variable, I ask the chatgpt for clearification and also went to the office hour, know the usage of the ifelse()function

## 3.Did you find any errors in the assignment?
Not yet

## 4.What did you find most useful about working with the tidy process?
First time, I have a feeling of we are finally working on real data and ask question and solve it.

## 5. What would you advise for someone starting out with tidy?
See the dataframe first to know what data we have and also what data range we have.
