---
title: "Assignment 5"
author: "Xiangling Chen"
date: "2024-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Call the Library
```{r load_libraries}
### load sf, tmap, and tidyverse
library(tidyverse)
library(sf)
library(tmap)
```

## Loading the Data
```{r load-data}
### load nc_schools data to read
###!!!! very important: no space in the ""
nc_schools <- read_sf("https://drive.google.com/uc?export=download&id=1hFPUSj5s-sApJKC5Vh5Ue5wpF_1gbClO")
glimpse(nc_schools)
```

```{r read_data}
### load nc_counties data to read
nc_counties <- read_sf("https://drive.google.com/uc?export=download&id=1hbaRC3hu6h9IMN4sNjY_BJhHyMTpMIT9")
glimpse(nc_counties)
```

## Our Study Area
```{r locate nc schools}
### Map the location
map1<-tm_shape(nc_schools) + tm_dots(col="red")
map2<- tm_shape(nc_counties) +tm_polygons("gray")
map2 + map1
```

# Analysis of Public Schools in Durham and Orange Counties, 2018-2022

## Selecting Our Data
```{r filter DO county}
### Select Counties
od_counties <-nc_counties|>
  filter(County == "Durham" | County == "Orange")
glimpse(od_counties)
```

```{r select DO schools}
### Select schools
od_schools <-nc_schools|>
  filter(county == "Orange"|county == "Durham")|>
  filter(sch_desg!="Charater")
glimpse(od_schools)
```

## Plot the Local Schools
```{r map-local school}
map2<-tm_shape(od_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(od_schools) + tm_dots(col="red")
map2+map1 + tm_layout(title="Orange and Durham County Schools")
```

```{r modify-map}
### Modified dots and add the title
map2<-tm_shape(od_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(od_schools) + tm_dots("bgn_grade", scale=4)
map2+map1+tm_layout(title="Orange and Durham County")
```

```{r calculate OD fail rate}
### Map local schools fail rate
map2<-tm_shape(od_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(od_schools) + tm_dots("pct_notprof_1819", scale=4)
map2+map1+ tm_layout(title="Orange & Durham Schools, % Fail 2018-2019")
```

## Computing Values
```{r computing_od_values}
### Compute num of studetns pass/fail at each school
od_schools <- od_schools |>
  #remove any null values
  drop_na(num_tested_1819, num_tested_2021,num_tested_2122)|>
  mutate(num_fail_1819 = round(num_tested_1819*pct_notprof_1819/100))|>
  mutate(num_fail_2021 = round(num_tested_2021*pct_notprof_2021/100))|>
  mutate(num_fail_2122 = round(num_tested_2122*pct_notprof_2122/100))|>
  mutate(num_pass_1819  = round(num_tested_1819 - num_fail_1819))|>
  mutate(num_pass_2021  = round(num_tested_2021 - num_fail_2021))|>
  mutate(num_pass_2122  = round(num_tested_2122 - num_fail_2122))
glimpse(od_schools)
```

```{r mutate od pass_rate}
od_schools<-od_schools|>
  mutate(total_pass = num_pass_1819 +num_pass_2021+num_pass_2122)|>
  mutate(total_tested = num_tested_1819+num_tested_2021+num_tested_2122)|>
  mutate(pct_pass = total_pass/total_tested*100)
glimpse(od_schools)
```

```{r map od study area}
#Let's map our study area
map1<-tm_shape(od_schools) + tm_dots("pct_pass", scale = 3)
map2<-tm_shape(od_counties) +tm_polygons(col = "gray")
map2+map1+tm_layout(title = "Orange & Durham Schools, % Pass")
```

```{r create od data frame}
# Create a data frame of values per county, including totals
od_schools_summary<- as.data.frame(od_schools)|>
  # Select just the columns we need
  select(county, total_tested, total_pass)|>
  #sum by county
  group_by(county)|>
  #this says to sum each column
  summarize(across(total_tested:total_pass, sum))|>
  #let's compute the pass pct for each county
  mutate(pass_pct = total_pass/total_tested*100)
glimpse(od_schools_summary)
```

## Yearly Trend
```{r od_yearly trend}
# create a data frame of values per county, including totals
od_schools_summary2<-as.data.frame(od_schools) |>
  #select just the columns we need
  select(county, num_tested_1819, num_tested_2021, num_tested_2122, num_pass_1819, num_pass_2021, num_pass_2122)|>
  #sum by county
  group_by(county)|>
  summarize(across(num_tested_1819:num_pass_2122,sum))|>
  #let's compute the pass pct for each year
  mutate(pass_pct_1819 = num_pass_1819/num_tested_1819*100)|>
  mutate(pass_pct_2021 = num_pass_2021/num_tested_2021*100)|>
  mutate(pass_pct_2122 = num_pass_2122/num_tested_2122*100)|>
  #and just return the columns that we need
  select(county, pass_pct_1819, pass_pct_2021, pass_pct_2122)
glimpse(od_schools_summary2)
```

```{r rearrange_od_data}
od_schools_summary2|>
  # wehave to rearrange our data so that it'll work with ggplot
  pivot_longer(pass_pct_1819:pass_pct_2122)|>
  #ggplot() is a plotting function
  ggplot()+geom_point(mapping = aes(x=name,y=value, color=county),size=3)
```

# Analysis of Public Schools in Camden and Gates Counties, 2018-2022
```{r filter cg_county}
cg_counties<-nc_counties|>
  filter(County == "Camden"|County =="Gates")
glimpse(cg_counties)
```

```{r select_cg_schools}
### select schools
cg_schools<-nc_schools|>
  filter(county == "Camden"|county =="Gates")|>
  filter(sch_desg!="Charater")
glimpse(cg_schools)
```

## Plot the Local Schools
```{r map_cg_local schools}
map2<-tm_shape(cg_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(cg_schools) + tm_dots(col="red")
map2+map1 + tm_layout(title="Camden and Gates County Schools")
```

```{r modify cg_map}
### Modified dots and add the title
map2<-tm_shape(cg_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(cg_schools) + tm_dots("bgn_grade", scale=4)
map2+map1+tm_layout(title="Camden and Gates County")
```

```{r map_cg_fail_rate}
### Map local schools fail rate
map2<-tm_shape(cg_counties) + tm_polygons(col = "grey")
map1 <-tm_shape(cg_schools) + tm_dots("pct_notprof_1819", scale=4)
map2+map1+ tm_layout(title="Camden & Gates Schools, % Fail 2018-2019")
```

## Computing Values
```{r computing_cg_values}
### Compute num of studetns pass/fail at each school
cg_schools <- cg_schools |>
  #remove any null values
  drop_na(num_tested_1819, num_tested_2021,num_tested_2122)|>
  mutate(num_fail_1819 = round(num_tested_1819*pct_notprof_1819/100))|>
  mutate(num_fail_2021 = round(num_tested_2021*pct_notprof_2021/100))|>
  mutate(num_fail_2122 = round(num_tested_2122*pct_notprof_2122/100))|>
  mutate(num_pass_1819  = round(num_tested_1819 - num_fail_1819))|>
  mutate(num_pass_2021  = round(num_tested_2021 - num_fail_2021))|>
  mutate(num_pass_2122  = round(num_tested_2122 - num_fail_2122))
glimpse(cg_schools)
```

```{r cg_pass_Rate}
cg_schools<-cg_schools|>
  mutate(total_pass = num_pass_1819 +num_pass_2021+num_pass_2122)|>
  mutate(total_tested = num_tested_1819+num_tested_2021+num_tested_2122)|>
  mutate(pct_pass = total_pass/total_tested*100)
glimpse(cg_schools)
```

```{r map_cg_study_area}
#Let's map our study area
map1<-tm_shape(cg_schools) + tm_dots("pct_pass", scale = 3)
map2<-tm_shape(cg_counties) +tm_polygons(col = "gray")
map2+map1+tm_layout(title = "Candem & Gates Schools, % Pass")
```

```{r create_cg_dataframe}
# Create a data frame of values per county, including totals
cg_schools_summary<- as.data.frame(cg_schools)|>
  # Select just the columns we need
  select(county, total_tested, total_pass)|>
  #sum by county
  group_by(county)|>
  #this says to sum each column
  summarize(across(total_tested:total_pass, sum))|>
  #let's compute the pass pct for each county
  mutate(pass_pct = total_pass/total_tested*100)
glimpse(cg_schools_summary)
```

```{r create_summary}
# create a data frame of values per county, including totals
cg_schools_summary2<-as.data.frame(cg_schools) |>
  #select just the columns we need
  select(county, num_tested_1819, num_tested_2021, num_tested_2122, num_pass_1819, num_pass_2021, num_pass_2122)|>
  #sum by county
  group_by(county)|>
  summarize(across(num_tested_1819:num_pass_2122,sum))|>
  #let's compute the pass pct for each year
  mutate(pass_pct_1819 = num_pass_1819/num_tested_1819*100)|>
  mutate(pass_pct_2021 = num_pass_2021/num_tested_2021*100)|>
  mutate(pass_pct_2122 = num_pass_2122/num_tested_2122*100)|>
  #and just return the columns that we need
  select(county, pass_pct_1819, pass_pct_2021, pass_pct_2122)
glimpse(cg_schools_summary2)
```

```{r map_cg_schools_rearranged}
cg_schools_summary2|>
  # we have to rearrange our data so that it'll work with ggplot
  pivot_longer(pass_pct_1819:pass_pct_2122)|>
  #ggplot() is a plotting function
  ggplot()+geom_point(mapping = aes(x=name,y=value, color=county),size=3)
```

#Reflection Questions
## 1. How long did the assignment take you to complete?
It takes me 3 hrs

## 2.Where did you get stuck? Where did you go for help?
I was stuck at understanding the ggplot, and I got help from the later class(the class on Wednesday)

## 3.Did you find any errors in the assignment?
Before I read the second part of filtering the county, I was kind of confused about the "&" used for filtering the Orange or Durham, but when I continue to read the HTML,Dr.Parr corrected it.

## 4.What did you find most useful about working with the tidy process?
I really like the all the process, especially the ggplot function. I like to see the visualization of the data.

## 5. What would you advise for someone starting out with tidy?
Be careful with the syntax and there are so many things you can play with. Not only the data but also the graphs.
